### 2015-06-18 chapters/ipa_meets_unicode.tex
update discussion tie bar

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The placement of diacritics
  \item The tie bar to indicate a close bond between sounds
\end{itemize}

@@

Third, for marking marking of voiceless pronunciation of voiced segments the IPA
uses the ring diacritic. Originally, the ring should be placed below the 
base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
However, in letters with long descenders the IPA also allows to put the ring 
above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
\uni{030A}. Yet, proper font design does not have any problem with rendering 
the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
propose to standardize on the ring below.

Finally, the tie bar to indicate affricates, doubly articulated consonants or
diphthongs can be placed either below or on top of the base characters, e.g.
<\charis{t͡s}> or <\charis{t͜s}>. For strict IPA encoding we propose to standardize on the tie bar
above the base characters, using Unicode \textsc{combining double inverted
breve} at \uni{0361}. Note that this Unicode character is placed between the two 
base characters to be combined. In principle, it is possible to combine more 
than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
necessary, we consider this possible, even though the rendering will never look 
good.

@@

>>>>

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The placement of diacritics
\end{itemize}

@@

Third, for marking marking of voiceless pronunciation of voiced segments the IPA
uses the ring diacritic. Originally, the ring should be placed below the 
base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
However, in letters with long descenders the IPA also allows to put the ring 
above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
\uni{030A}. Yet, proper font design does not have any problem with rendering 
the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
propose to standardize on the ring below. As a principle, for strict IPA 
encoding only one options should be allowed.

% ==========================
\section{Pitfall: Tie bar}
\label{pitfall-tie-bar}
% ==========================

The IPA includes a tie bar to indicate affricates, doubly articulated consonants
or diphthongs. This character can be placed above or below the base characters,
e.g. <\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice
between the two symbols is purely for legible rendering; there is no difference
in semantics between the two symbols. The tie-bar is a convenient diacritic for
visually tokenizing input strings into chunks of phonetically salient groups.
However, rendering is such a problematic issue for tie bars in general that many
linguists simply do not use them. Just looking at a few different fonts already
indicates that actually no font designer really gets the placement right in
combination with superscripts and subscripts. If really necessary, we propose to
standardize on the tie bar above the base characters, using Unicode
\textsc{combining double inverted breve} at \uni{0361}.

\begin{itemize}
  \item[] {\fontspec{Times New Roman}Times new Roman: <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] {\small \fontspec{CharisSIL}CharisSIL: <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] {\footnotesize \fontspec{Monaco}Monaco: <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] {\fontspec{DoulosSIL}DoulosSIL: <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] Linux Libertine: <t̥ʰ͡s> <t̥ʰ͜s>
\end{itemize}

Tie bars are a special type of character in the sense that they do not belong to
a segment, but bind two graphemes together. This actually turns out to be rather
different from the Unicode conceptions, see
Section~\ref{pitfall-different-notions-of-diacritics}. According to the spirit
of the IPA, it is possible to combine more than two base characters into one tie
bar, but this is not possible with Unicode (i.e. there is no possibility to draw
a tie bar over three of four characters). It is possible to indicate such larger
groups by repeating the tie bar, like in <a͡o͡u>. If really necessary, we
consider this possible, even though the rendering will never look good.

Most importantly though, the tie-bar is unnecessary for computational
tokenization given the design of orthography profiles, as described in
Section~\ref{usecase-tokenization}. The principle idea of such tokenization is 
that groups of characters that belong together are defined in an orthography 
profile, making it superfluous to indicate them explicitly by tie bars. For this 
reason we actually would like to suggest to remove the tie bar from any 
``strict'' IPA encoding.

@@

% problem with tie bar!!!

In Section~\ref{pitfall-different-notions-of-diacritics}, we highlight
the division of linguistic diacritics into two categories in the Unicode
Standard: Spacing Modifier Letters and Combining Diacritical Marks.\footnote{The
former contain characters like aspiration < ʰ > which is user specified (before
or after the base character) and the latter which combines with the base
character.} The tie-bar is somewhere in between these two character classes,
i.e.~when the combining breve (above or below) is told to combine within a
sequence of Unicode characters, it will bind to the character immediately
preceding it -- the so-called base character -- regardless if that base
character is a Spacing Modifier Letter \textit{or} Combining Diacritical Mark.
\footnote{In some cases conversion of a string sequence into Unicode
Normalization Form C alleviates the problem by composing sequences of characters
like <e> + < ́ > into single character < é > (see however diacritics like the
rhotic hook in ).}
